---
title: 'Weekly Exercises #3'
author: "Alex Ang"
output: 
  html_document:
    keep_md: TRUE
    toc: TRUE
    toc_float: TRUE
    df_print: paged
    code_download: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error=TRUE, message=FALSE, warning=FALSE)
```

```{r libraries}
library(tidyverse)     # for graphing and data cleaning
library(gardenR)       # for Lisa's garden data
library(lubridate)     # for date manipulation
library(ggthemes)      # for even more plotting themes
library(geofacet)      # for special faceting with US map layout
theme_set(theme_minimal())       # My favorite ggplot() theme :)
```

```{r data}
# Lisa's garden data
data("garden_harvest")

# Seeds/plants (and other garden supply) costs
data("garden_spending")

# Planting dates and locations
data("garden_planting")

# Tidy Tuesday data
kids <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-09-15/kids.csv')
```

## Warm-up exercises with garden data

These exercises will reiterate what you learned in the "Expanding the data wrangling toolkit" tutorial. If you haven't gone through the tutorial yet, you should do that first.

  1. Summarize the `garden_harvest` data to find the total harvest weight in pounds for each vegetable and day of week (HINT: use the `wday()` function from `lubridate`). Display the results so that the vegetables are rows but the days of the week are columns.

```{r}
garden_harvest %>%
  group_by(vegetable, day_of_week=wday(date, label=TRUE)) %>%
  summarize(weight_lbs=sum(weight)*0.00220462)
```

  2. Summarize the `garden_harvest` data to find the total harvest in pound for each vegetable variety and then try adding the plot from the `garden_planting` table. This will not turn out perfectly. What is the problem? How might you fix it?
  
We have more rows that we originally had. We can fix this by getting rid of the variable plot because it is creating duplicates when we merge the datasets. 

```{r}
data("garden_planting")

garden_harvest %>%
  group_by(vegetable, variety) %>% 
  summarize(weight_lbs=sum(weight)*0.00220462) %>%
  left_join(garden_planting, 
            by= c("vegetable", "variety"))
  
```

  3. I would like to understand how much money I "saved" by gardening, for each vegetable type. Describe how I could use the `garden_harvest` and `garden_spending` datasets, along with data from somewhere like [this](https://products.wholefoodsmarket.com/search?sort=relevance&store=10542) to answer this question. You can answer this in words, referencing various join functions. You don't need R code but could provide some if it's helpful.

```{r}
data("garden_spending")
```

I would begin by merging the garden harvest and garden spending datasets so we can look at the weight of the harvests and compare it to the the price of our vegetables. I would then group by vegetable and find the total weight of each vegetable harvested and total price without tax for each vegetable. I would then compare it with the Whole Food prices by looking at their price per pound and multiplying it by how ever many pounds of the same vegetable that Lisa harvested. Lastly, I would find how much was saved by subtracting this number by the total price without tax for that same vegetable to see how much was saved compared to the Whole Foods price. 

  4. Subset the data to tomatoes. Reorder the tomato varieties from smallest to largest first harvest date. Create a barplot of total harvest in pounds for each variety, in the new order.

```{r}
garden_harvest %>%
  filter(vegetable %in% "tomatoes") %>%
  group_by(variety) %>%
  summarize(weight_lbs=sum(weight*0.00220462), first_harvest = min(date))  %>%
  ggplot(aes(y=fct_reorder(variety, first_harvest), x=weight_lbs)) +
  geom_col() + 
  labs(y="", x="weight(lbs)", title="Variety of Tomatoes by First Harvest date")
```

  5. In the `garden_harvest` data, create two new variables: one that makes the varieties lowercase and another that finds the length of the variety name. Arrange the data by vegetable and length of variety name (smallest to largest), with one row for each vegetable variety. HINT: use `str_to_lower()`, `str_length()`, and `distinct()`.
  
```{r}
garden_harvest %>%
  mutate(var_lower=str_to_lower(variety), var_length=str_length(variety)) %>%
  arrange(vegetable, (fct_infreq(variety))) %>%
  distinct(vegetable, var_length)
  
```

  6. In the `garden_harvest` data, find all distinct vegetable varieties that have "er" or "ar" in their name. HINT: `str_detect()` with an "or" statement (use the | for "or") and `distinct()`.
  

```{r}
garden_harvest %>%
  mutate(var=str_detect(variety, "er|ar")) %>%
  filter(var==TRUE) %>%
  distinct(variety)
```


## Bicycle-Use Patterns

In this activity, you'll examine some factors that may influence the use of bicycles in a bike-renting program.  The data come from Washington, DC and cover the last quarter of 2014.

Two data tables are available:

- `Trips` contains records of individual rentals
- `Stations` gives the locations of the bike rental stations

Here is the code to read in the data. We do this a little differently than usualy, which is why it is included here rather than at the top of this file. To avoid repeatedly re-reading the files, start the data import chunk with `{r cache = TRUE}` rather than the usual `{r}`.

```{r cache=TRUE}
data_site <- 
  "https://www.macalester.edu/~dshuman1/data/112/2014-Q4-Trips-History-Data.rds" 
Trips <- readRDS(gzcon(url(data_site)))
Stations<-read_csv("http://www.macalester.edu/~dshuman1/data/112/DC-Stations.csv")
```

**NOTE:** The `Trips` data table is a random subset of 10,000 trips from the full quarterly data. Start with this small data table to develop your analysis commands. **When you have this working well, you should access the full data set of more than 600,000 events by removing `-Small` from the name of the `data_site`.**

### Temporal patterns

It's natural to expect that bikes are rented more at some times of day, some days of the week, some months of the year than others. The variable `sdate` gives the time (including the date) that the rental started. Make the following plots and interpret them:

  7. A density plot, which is a smoothed out histogram, of the events versus `sdate`. Use `geom_density()`.
  
```{r}
Trips %>%
  ggplot(aes(x=sdate)) + 
  geom_density() + 
  labs(y="", x="", title="Distribution of When Rentals Were Started")
```
  
  8. A density plot of the events versus time of day.  You can use `mutate()` with `lubridate`'s  `hour()` and `minute()` functions to extract the hour of the day and minute within the hour from `sdate`. Hint: A minute is 1/60 of an hour, so create a variable where 3:30 is 3.5 and 3:45 is 3.75.
  
```{r}
Trips %>%
  mutate(time_of_day=hour(sdate) + minute(sdate)/60) %>%
  ggplot(aes(x=time_of_day)) + 
  geom_density() + 
  labs(y="", x="", title="Distribution of Rentals Over Time of Day")
```
  
  9. A bar graph of the events versus day of the week. Put day on the y-axis.
  
```{r}
Trips %>%
  mutate(dow=wday(sdate, label=TRUE)) %>%
  ggplot(aes(y=dow)) +
  geom_bar() + 
  labs(title="Events versus day of the week", y="", x="")
  
```
  
  10. Facet your graph from exercise 8. by day of the week. Is there a pattern?
  
There is a clear pattern, where the weekdays experience a drop in the middle and then come back up. On the weekends, the shape of the density plot is similar for Saturday and Sunday but set itself apart from the rest of the weekdays. 
  
```{r}
Trips %>%
  mutate(time_of_day=hour(sdate) + minute(sdate)/60) %>%
  mutate(dow=wday(sdate, label=TRUE)) %>%
  ggplot(aes(x=time_of_day)) + 
  geom_density() + 
  facet_grid(.~dow) +
  labs(y="", x="", title="Distribution of Rentals @ Different Times of Each Day of the Week")
```
  
The variable `client` describes whether the renter is a regular user (level `Registered`) or has not joined the bike-rental organization (`Causal`). The next set of exercises investigate whether these two different categories of users show different rental behavior and how `client` interacts with the patterns you found in the previous exercises. 

  11. Change the graph from exercise 10 to set the `fill` aesthetic for `geom_density()` to the `client` variable. You should also set `alpha = .5` for transparency and `color=NA` to suppress the outline of the density function.
  
```{r}
Trips %>%
  mutate(time_of_day=hour(sdate) + minute(sdate)/60) %>%
  mutate(dow=weekdays(sdate)) %>%
  ggplot(aes(x=time_of_day, fill=client)) + 
  geom_density(alpha=.5, color=NA) + 
  facet_grid(.~dow) + 
  labs(y="", x="", title="Distribution of Rentals @ Different Times of Each Day of the Week")
```

  12. Change the previous graph by adding the argument `position = position_stack()` to `geom_density()`. In your opinion, is this better or worse in terms of telling a story? What are the advantages/disadvantages of each?
  
I think using position_stack is better because we can see it in the context of each of the two variables. One advantage is that it allows us to see which group is most frequent for a given value, however it makes it hard to understand the distribution of a group that is not on the bottom of the chart. 
  
```{r}
Trips %>%
  mutate(time_of_day=hour(sdate) + minute(sdate)/60) %>%
  mutate(dow=weekdays(sdate)) %>%
  ggplot(aes(x=time_of_day, fill=client)) + 
  geom_density(alpha=.5, color=NA, position=position_stack())+ 
  facet_grid(.~dow) + 
  labs(y="", x="", title="Distribution of Rentals @ Different Times of Each Day of the Week")
```
  
  13. In this graph, go back to using the regular density plot (without `position = position_stack()`). Add a new variable to the dataset called `weekend` which will be "weekend" if the day is Saturday or Sunday and  "weekday" otherwise (HINT: use the `ifelse()` function and the `wday()` function from `lubridate`). Then, update the graph from the previous problem by faceting on the new `weekend` variable. 
  
  
```{r}
Trips %>%
  mutate(time_of_day=hour(sdate) + minute(sdate)/60,
         days_of_week=wday(sdate),
         weekend=ifelse(days_of_week>5, "weekend", "weekday")) %>%
  ggplot(aes(x=time_of_day, fill=client)) + 
  geom_density(alpha=.5, color=NA)+ 
  facet_grid(.~weekend) + 
  labs(y="", x="", title="Distribution of Rentals @ Different Times of Weekdays vs Weekends")
```
  
  14. Change the graph from the previous problem to facet on `client` and fill with `weekday`. What information does this graph tell you that the previous didn't? Is one graph better than the other?
  
```{r}
Trips %>%
  mutate(time_of_day=hour(sdate) + minute(sdate)/60,
         days_of_week=wday(sdate),
         weekend=ifelse(days_of_week>5, "weekend", "weekday")) %>%
  ggplot(aes(x=time_of_day, fill=weekend)) + 
  geom_density(alpha=.5, color=NA)+ 
  facet_grid(.~client) + 
  labs(y="", x="", title="Distribution of Rentals @ Different Times for Casual vs Registered Riders")
```
  
### Spatial patterns

  15. Use the latitude and longitude variables in `Stations` to make a visualization of the total number of departures from each station in the `Trips` data. Use either color or size to show the variation in number of departures. We will improve this plot next week when we learn about maps!
  
  
```{r}
Trips %>% 
  group_by(sstation) %>%
  summarize(n_trips=n()) %>%
  left_join(Stations, by=c("sstation"="name")) %>%
  ggplot(aes(y=lat, x=long, size=n_trips)) +
  geom_point(alpha=0.3) + 
  labs(size="# of trips", y="latitude", x="longitude")

```

  16. Only 14.4% of the trips in our data are carried out by casual users. Create a plot that shows which area(s) have stations with a much higher percentage of departures by casual users. What patterns do you notice? (Again, we'll improve this next week when we learn about maps).
  
```{r}
Trips %>% 
  mutate(client2= client=="Casual") %>%
  group_by(sstation) %>%
  summarize(cas_percentage=sum(client2)/n()) %>%
  left_join(Stations, by=c("sstation"="name")) %>%
  ggplot(aes(y=lat, x=long, size=cas_percentage)) +
  geom_point(alpha=0.3) + 
  labs(size="percentage of riders", y="latitude", x="longitude")
```
  
### Spatiotemporal patterns

  17. Make a table with the ten station-date combinations (e.g., 14th & V St., 2014-10-14) with the highest number of departures, sorted from most departures to fewest. Save this to a new dataset and print out the dataset. Hint: `as_date(sdate)` converts `sdate` from date-time format to date format. 
  
```{r}
Trips %>%
  mutate(date=as_date(sdate)) %>%
  group_by(sstation, date) %>%
  summarize(departures=n()) %>%
  arrange(desc(departures)) %>%
  head(10) 
```
  
  18. Use a join operation to make a table with only those trips whose departures match those top ten station-date combinations from the previous part.
  
```{r}
top_ten <- Trips %>%
  mutate(date=as_date(sdate)) %>%
  group_by(sstation, date) %>%
  summarize(departures=n()) %>%
  arrange(desc(departures)) %>%
  head(10) 

Trips %>%
  mutate(date=as_date(sdate)) %>%
  inner_join(top_ten, by=c("sstation", "date"))
```
  
  19. Build on the code from the previous problem (ie. copy that code below and then %>% into the next step.) and group the trips by client type and day of the week (use the name, not the number). Find the proportion of trips by day within each client type (ie. the proportions for all 7 days within each client type add up to 1). Display your results so day of week is a column and there is a column for each client type. Interpret your results.
  
```{r}
Trips %>%
  mutate(date=as_date(sdate)) %>%
  inner_join(top_ten, by = c("sstation", "date")) %>%
  mutate(day_of_week=wday(date, label=TRUE)) %>%
  group_by(day_of_week, client) %>%
  summarize(n_trips=n()) %>%
  group_by(client) %>% 
  mutate(prop=n_trips/sum(n_trips)) %>%
  pivot_wider(id_cols=day_of_week, names_from=client, values_from=prop)
```

## GitHub link

  20. Below, provide a link to your GitHub page with this set of Weekly Exercises. Specifically, if the name of the file is 03_exercises.Rmd, provide a link to the 03_exercises.md file, which is the one that will be most readable on GitHub.


**DID YOU REMEMBER TO UNCOMMENT THE OPTIONS AT THE TOP?**
